
## Agricultural Stress Index (ASI) Analysis 

Syria 2024/2025 Drought Season

This section analyzes the evolution and spatial distribution of the Agricultural Stress Index (ASI) across Syrian governorates during the critical 2024/2025 winter cropping season (November 2024 to April 2025), a period highlighted in FAO’s drought alert as the most severe in 36 years.

We use ASI data obtained through the `cumulus` R package via FAO GIEWS, filtered by province and dekad to explore stress trends across crop-growing areas.

### Load and Filter ASI Data



```{r}
box::use(
  cumulus
)
box::use(
  lubridate[...],
  dplyr[...],
  tidyr[...],
  readr[...],
  ggplot2[...],
  glue[...],
  forcats[...],
  gghdx[...]
)
df_asi <- cumulus$fao_asi_adm1_tabular(iso3 = "syr")

gghdx()
```

```{r}
df_filtered <- df_asi |> 
  mutate(
    mo_num = as.numeric(Month),
       harvest_year = case_when(
      mo_num %in% c(11, 12) ~ Year + 1,
      TRUE ~ Year
    ),
    mo_chr = fct_expand(month.abb[mo_num],month.abb),
    date = glue("{Year}-{Month}-21")
  ) |> 
  filter(
    Dekad == 3,
    mo_num %in% c(11,12,1,2,3,4)
  )


df_filtered |> 
  mutate(
    mo_fct= fct_relevel(mo_chr, c("Nov","Dec","Jan","Feb","Mar","Apr")),
    harvest_year_label = case_when(
      harvest_year %in% c(2023:2025)~as.character(harvest_year),
      .default =  "Other"
    ),
        alpha_flag = if_else(harvest_year_label == "Other", "low", "high")  # for alpha mapping

  ) |> 
  ggplot(
    aes(x= mo_fct,
        y= Data,
        group = harvest_year,
        color =harvest_year_label,
        alpha = alpha_flag)
  )+
   scale_alpha_manual(values = c("low" = 0.3, "high" = 1)) +
  # scale_alpha_discrete(values = c("other"=0.1))+
  # geom_point()+
  geom_line()+
  facet_wrap(~Province)+
  theme(
    axis.text.x = element_text(angle = 90)
  )

```


